To Batch or Not to Batch: Test-Ordering Variability in the ED

### 2024-09-09 manuscript/manuscript-v32.tex
getting going

\hyperref[table:first_stage]{Table 1} illustrates the ``first stage''
results from both linear and probit models, emphasizing the robust
relevance of the instrument across different statistical frameworks. In
both modeling approaches, a 10 percentage point increase in the
batch-tendency of physicians consistently results in a significant
increase in the likelihood of batch ordering tests in the emergency
department---1.15 percentage points in the linear model and 93.24
percentage points in the probit model. This consistent significance
across models underscores the strength and relevance of batch tendency
as an instrumental variable. Given the robustness to different model
specifications, we move forward with a probit model in our first stage
to estimate the effect of batch ordering on patient outcomes.

@@

This reduced form will allow us to check the strength of our instrument.
To study the effects of test batching in the ED on an outcome \(Y_i\),
we employ a two-stage estimation approach using our baseline sample:

@@

\begin{equation}
P(Batched_i = 1) = \Phi(\delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i)
\end{equation}

Where \(Y_i\) represents our main outcomes of interest: length of stay,
72 hour readmission, and resource utilization, and \(X_i\) is the same
as in the reduced-form approach. \(Batched_i\) variable suffers from
potential endogeneity concerns. For example, injury severity may be
unobserved and correlated with need to run multiple tests, which in turn
also affects length of stay. Hence, we instrument \(Batched_i\) with the
assigned physician \(j\)'s underlying tendency to batch,
\(Tendency_{i,j}^{phys}\). We cluster robust standard errors at the
physician level to account for the assignment process of patients to
physicians.

>>>>

\hyperref[table:first_stage]{Table 1} illustrates the ``first stage''
results from both linear and probit models, emphasizing the robust
relevance of the instrument across different statistical frameworks. In
both modeling approaches, a 10 percentage point increase in the
batch-tendency of physicians consistently results in a significant
increase in the likelihood of batch ordering tests in the emergency
department---1.15 percentage points in the linear model and 93.24
percentage points in the probit model. This consistent significance
across models underscores the strength and validity of the
batch-tendency as an instrumental variable, confirming its utility in
controlling for unobserved confounders and providing a reliable basis
for further causal inference. Clustering standard errors at the provider
level ensures robust inference, reinforcing the instrument's credibility
irrespective of the analytical method employed.

@@

This reduced form will allow us to check that our instrument is a strong
instrument. To study the effects of test batching in the ED on an
outcome \(Y_i\), we estimate the following 2SLS equations using our
baseline sample:

@@

\begin{equation}
Batched_i = \delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i + \nu_i
\end{equation}

Where \(Y_i\) represents our main outcomes of interest: length of stay,
72 hour readmission, and resource utilization. and \(X_i\) is the same
as in the reduced-form approach. \(Batched_i\) variable suffers from
potential endogeneity concerns. For example, injury severity may be
unobserved and correlated with need to run multiple tests, which in turn
also affects length of stay. Hence, we instrument \(Batched_i\) with the
assigned physician \(j\)'s underlying tendency to batch,
\(Tendency_{i,j}^{phys}\). We cluster robust standard errors at the
physician level to account for the assignment process of patients to
physicians.
